\LARGE \bf Mean Estimation from One-Bit Measurements

### 2021-09-30 uniform-are-weirdos.tex
Commented on O(1/n) in Prop. 1

To prove inequality~\eqref{eqn:uniforms-are-easy}, we proceed in two steps,
both quite similar.
First, we define an initial estimator $\theta\init_n$.
Let $\epsilon > 0$, which we will determine presently, though we will
take $n \epsilon \to \infty$ as $n \to \infty$, so that we may assume
w.l.o.g.\ that $\theta \in [-n\epsilon/2, n \epsilon/2]$. Take the interval
$[-n\epsilon, n\epsilon]$, and construct
$m$ thresholds at intervals of size $2 n \epsilon / m$; let
the $j$th such threshold be
\begin{equation*}
  t_j \defeq -n \epsilon + \frac{2 n (j-1) \epsilon}{m}
\end{equation*}
Then we ``assign'' observations to each pair of thresholds, so that
threshold $j$ corresponds to observations $I_j \defeq \{\frac{n (j - 1)}{m}
+ 1, \ldots, \frac{n j}{m}\}$, of which there are $n/m$. For each index $i
\in I_j$, we set
\begin{equation*}
  B_i = \begin{cases}
    1 & \mbox{if~} X_i \ge t_j \\
    0 & \mbox{otherwise}.
  \end{cases}
\end{equation*}
Then we simply set $\theta_n\init$ to be the \newtext{minimal} threshold for which
$B_i = 0$ for all observations $X_i$ corresponding to that threshold.

Let us now consider the probability that $\theta_n\init$ is substantially
wrong. For notational simplicity, let $U_i = (1 + X_i - \theta) / 2$, so that
the $U_i$ are uniform on $[0, 1]$.
First, note that we always have $\theta_n\init \ge \theta
- \frac{2 n \epsilon}{m}$, because no observations will be below
the appropriate threshold. Let $j\opt$ be the \newtext{largest} index $j$
for which $\theta \newtext{\ge} t_{j\opt}$, and consider
the index sets $I_{j\opt}$, $I_{j\opt + 1}$, and so on.
The event $\theta\init_n \ge t_{j\opt} + \frac{2 n \epsilon}{m}$ may
occur only if for each of the $n/m$ observations in the set
$I_{j\opt + 1}$, we have $U_i \ge \frac{n \epsilon}{m}$. Thus,
\begin{align*}
  P_\theta\left(\theta_n\init \ge \theta + \frac{2 n \epsilon}{m}
  \right) \le
  P_\theta\left(\theta_n\init \ge t_{j\opt} + \frac{2 n \epsilon}{m}
  \right)
  \le \left(1 - \frac{n \epsilon}{m}\right)^\frac{n}{m}
  \le \exp\left(-\frac{n^2\epsilon}{m^2}\right).
\end{align*}
Setting the number of bins $m = \sqrt{n}$,
the resolution $\epsilon = 2 \log n / n$, we obtain
$P_\theta(\theta_n\init \ge t_{j\opt} + 4 \log n / \sqrt{n})
\le n^{-2}$. Thus we have
\begin{align}
  \label{eqn:quality-of-initial-estimate}
  \sup_{|\theta| \le \log n} P_\theta\left(|\theta_n\init - \theta| \ge
  \frac{8 \log n}{\sqrt{n}}\right) \le \frac{1}{n^2}.
\end{align}

The second stage estimator follows roughly the same strategy, except that
the resolution of the bins is tighter. In particular, let us assume that
$|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$, which happens
eventually by inequality~\eqref{eqn:quality-of-initial-estimate}.  (We will
assume this tacitly for the remainder of the argument.)  Consider the
interval $\Theta_n \defeq \theta_n\init + [-\frac{16 \log n}{\sqrt{n}},
  \frac{16 \log n}{\sqrt{n}}]$ centered at $\theta_n\init$; we know that the
interval includes $[\theta - \frac{8 \log n}{\sqrt{n}}, \theta + \frac{8
    \log n}{\sqrt{n}}]$. Without loss of generality we assume $\theta_n\init
= 0$.  Following precisely the same discretization strategy as that for
$\theta_n\init$, we divide $\Theta_n$ into $m$ equal intervals, with
thresholds $t_j = -\frac{16 \log n}{\sqrt{n}} + \frac{32 (j - 1) \log n}{m
  \sqrt{n}}$; let $\epsilon_n = \frac{32 \log n}{m \sqrt{n}}$ be the width of
these intervals.  Then following exactly the same reasoning as above, we
assign indices $I_j = \{\frac{n(j - 1)}{m} + 1, \ldots, \frac{n j}{m}\}$ and
for $i \in I_j$, set $B_i = 1$ if $X_i \ge t_j$. We define $\theta_n$ to be
the \newtext{minimal} threshold $t_j$ for which $B_i = 0$ for all observations $X_i
\in I_j$. Then following precisely the reasoning above, we have (on the
event that $|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$)
\begin{equation*}
  P_\theta(|\theta_n - \theta|
  \ge 2 \epsilon_n)
  \le (1 - \epsilon_n)^\frac{n}{m}
  \le \exp\left(-\frac{n \epsilon_n}{m}\right)
  = \exp\left(-\frac{32 \sqrt{n} \log n}{m^2}\right).
\end{equation*}
Set $m = 4 n^{1/4}$ to obtain the claimed
result~\eqref{eqn:uniforms-are-easy}.

@@

%
Denote by $D^\star$ the optimal MSE in the Gaussian CEO with $L$ observers and under a total sum-rate $r = r_1 + \ldots +r_L$. An expression for $D^\star$ as a function of $r$ is give as \cite[Eq. 10]{chen2004upper}:
\begin{equation} \label{eq:ceo_optimal_sumrate}
r = \frac{1}{2} \log^+ \left[ \frac{\sigma_\theta^2}{D^\star} \left( \frac{D^\star L}{ D^\star L - \sigma^2 + D^\star \sigma^2 / \sigma_\theta^2 }\right)^L  \right].
\end{equation}
For the special case where $r = n$ and $L=n$, we have
\begin{equation} \label{eq:ceo_optimal_sumrate2}
n = \frac{1}{2} \log_2 \left[ \frac{\sigma_\theta^2}{D^\star} \left(\frac{ D^\star n }{D^\star n - \sigma^2 + D^\star \sigma^2/\sigma_\theta^2 }  \right)^n  \right].
\end{equation}
Consider the distributed encoding setting (iii) in the case where $f(x) = \Ncal(0,\sigma^2)$ and the prior on $\Theta$ is $\pi = \Ncal(0,\sigma_\theta^2)$. The Gaussian CEO problem of \cite{viswanathan1997quadratic} with a unit bitrate $r_1=\ldots = r_n =1$ at each terminal and blocklength $k=1$ reduces to our distributed setting (iii). Since $D^\star$ satisfying \eqref{eq:ceo_optimal_sumrate2} describes the MSE in the CEO setting under an optimal allocation of the sum-rate $r = n$ among $n$ encoders, it provides a lower bound to the minimal MSE in estimating $\theta$ in the distributed setting. By considering the limit $n\rightarrow \infty$ in \eqref{eq:ceo_optimal_sumrate2}, we see that 
\[
D^\star = \frac{ 4\sigma^2 }{3n + 4 \sigma^2 / \sigma_\theta^2 } + o(n^{-1}) =  \frac{4\sigma^2}{3n} + o(n^{-1}). 
\]
This implies Proposition~\ref{prop:CEO}. 
%We note that although the lower bound \eqref{eq:ceo_bound} was derived assuming the optimal allocation of $n$ bits per observation among the encoders, this bound cannot be tightened by considering the MSE in the CEO setting while enforcing the condition $r_1=\ldots = r_n = 1$. Indeed, an upper bound for the CEO MSE under the condition $r_1=\ldots = r_n = 1$ follows from \cite{KipnisRini2019}, and leads to
%\[
%D^\star \leq  \left( \frac{1}{\sigma_\theta^2} +  \frac{3n}{4\sigma^2 + \sigma_\theta^2} \right)^{-1}   =
%\frac{4 \sigma^2}{3n} +  \frac{\sigma_\theta^2}{3n} + O(n^{-2}),
%\]
%which is equivalent to \eqref{eq:ceo_bound} when $\sigma_\theta$ is small.

>>>>

To prove inequality~\eqref{eqn:uniforms-are-easy}, we proceed in two steps,
both quite similar.
First, we define an initial estimator $\theta\init_n$.
Let $\epsilon > 0$, which we will determine presently, though we will
take $n \epsilon \to \infty$ as $n \to \infty$, so that we may assume
w.l.o.g.\ that $\theta \in [-n\epsilon/2, n \epsilon/2]$. Take the interval
$[-n\epsilon, n\epsilon]$, and construct
$m$ thresholds at intervals of size $2 n \epsilon / m$; let
the $j$th such threshold be
\begin{equation*}
  t_j \defeq -n \epsilon + \frac{2 n (j-1) \epsilon}{m}
\end{equation*}
Then we ``assign'' observations to each pair of thresholds, so that
threshold $j$ corresponds to observations $I_j \defeq \{\frac{n (j - 1)}{m}
+ 1, \ldots, \frac{n j}{m}\}$, of which there are $n/m$. For each index $i
\in I_j$, we set
\begin{equation*}
  B_i = \begin{cases}
    1 & \mbox{if~} X_i \newtext{- 1} \ge t_j \\
    0 & \mbox{otherwise}.
  \end{cases}
\end{equation*}
Then we simply set $\theta\init_n$ to be the \newtext{minimal} threshold for which
$B_i = 0$ for all observations $X_i$ corresponding to that threshold. \newtext{Denote by $j^*$ the index of the threshold corresponding to $\theta\init_n$.}

\newtext{Let us now consider the probability that $\theta_n\init$ is substantially wrong. Set $\theta_M \equiv \max_{i \in I_j^*} X_i - 1$. Note that we always have $\theta\init_n \ge \theta_M$ because no observations will be above
$t_j^*+1$, and that $\theta\init_n \le \theta_M + 2n\epsilon/m$. 
 In addition, 
\begin{align*}
P_\theta\left( \left| \theta_M - \theta \right| \ge \frac{2n\epsilon}{m} \right) = \left(1 - \frac{2 n \epsilon}{m} \right)^{n/m}.
\end{align*}
Putting it all together using the triangle inequality, we have
\begin{align*}
  P_\theta\left( \left| \theta\init_n - \theta \right| \ge \frac{4n\epsilon}{m} \right) \le \left(1 - \frac{2 n \epsilon}{m} \right)^{n/m} \le e^{-2 \frac{n^2}{m^2} \epsilon}
  .
\end{align*}
%For notational simplicity, let $U_i = (1 + X_i - \theta) / 2$, so that the $U_i$ are uniform on $[0, 1]$. First, note that we always have $\theta_n\init \ge \theta - \frac{2 n \epsilon}{m}$, because no observations will be below
% the appropriate threshold. Let $j\opt$ be the \newtext{largest} index $j$
% for which $\theta \newtext{\ge} t_{j\opt}$, and consider
%the index sets $I_{j\opt}$, $I_{j\opt + 1}$, and so on.
%The event $\theta\init_n \ge t_{j\opt} + \frac{2 n \epsilon}{m}$ may
%occur only if for each of the $n/m$ observations in the set
%$I_{j\opt + 1}$, we have $U_i \ge \frac{n \epsilon}{m}$. Thus,
%\begin{align*}
%  P_\theta\left(\theta_n\init \ge \theta + \frac{2 n \epsilon}{m}
%  \right) \le
%  P_\theta\left(\theta_n\init \ge t_{j\opt} + \frac{2 n \epsilon}{m}
%  \right)
%  \le \left(1 - \frac{n \epsilon}{m}\right)^\frac{n}{m}
%  \le \exp\left(-\frac{n^2\epsilon}{m^2}\right).
%\end{align*}
Therefore, setting the number of bins $m = \sqrt{n}$ and
the resolution $\epsilon = 2 \log n / n$, 
\begin{align}
  \label{eqn:quality-of-initial-estimate}
  \sup_{|\theta| \le \log n} P_\theta\left(|\theta_n\init - \theta| \ge
  \frac{8 \log n}{\sqrt{n}}\right) \le \frac{1}{n^4}.
\end{align}
}
%$P_\theta(\theta_n\init \ge t_{j\opt} + 4 \log n / \sqrt{n})
%\le n^{-2}$. Thus we have
%\begin{align}
%  \label{eqn:quality-of-initial-estimate}
%  \sup_{|\theta| \le \log n} P_\theta\left(|\theta_n\init - \theta| \ge
%  \frac{8 \log n}{\sqrt{n}}\right) \le \frac{1}{n^2}.
%\end{align}

The second stage estimator follows roughly the same strategy, except that
the resolution of the bins is tighter. In particular, let us assume that
$|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$, which happens
eventually by inequality~\eqref{eqn:quality-of-initial-estimate}.  (We will
assume this tacitly for the remainder of the argument.)  Consider the
interval $\Theta_n \defeq \theta_n\init + [-\frac{16 \log n}{\sqrt{n}},
  \frac{16 \log n}{\sqrt{n}}]$ centered at $\theta_n\init$; we know that the
interval includes $[\theta - \frac{8 \log n}{\sqrt{n}}, \theta + \frac{8
    \log n}{\sqrt{n}}]$. Without loss of generality we assume $\theta_n\init
= 0$.  Following precisely the same discretization strategy as that for
$\theta_n\init$, we divide $\Theta_n$ into $m$ equal intervals, with
thresholds $t_j = -\frac{16 \log n}{\sqrt{n}} + \frac{32 (j - 1) \log n}{m
  \sqrt{n}}$; let $\epsilon_n = \frac{32 \log n}{m \sqrt{n}}$ be the width of
these intervals.  Then following exactly the same reasoning as above, we
assign indices $I_j = \{\frac{n(j - 1)}{m} + 1, \ldots, \frac{n j}{m}\}$ and
for $i \in I_j$, set $B_i = 1$ if $X_i \newtext{- 1} \ge t_j$. We define $\theta_n$ to be
the \newtext{minimal} threshold $t_j$ for which $B_i = 0$ for all observations $X_i
\in I_j$. Then following precisely the reasoning above, we have (on the
event that $|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$)
\begin{equation*}
  P_\theta(|\theta_n - \theta|
  \ge 2 \epsilon_n)
  \le (1 - \epsilon_n)^\frac{n}{m}
  \le \exp\left(-\frac{n \epsilon_n}{m}\right)
  = \exp\left(-\frac{32 \sqrt{n} \log n}{m^2}\right).
\end{equation*}
Set $m = 4 n^{1/4}$ to obtain the claimed
result~\eqref{eqn:uniforms-are-easy}.

@@

%
Denote by $D^\star$ the optimal MSE in the Gaussian CEO with $L$ observers and under a total sum-rate $r = r_1 + \ldots +r_L$. An expression for $D^\star$ as a function of $r$ is give as \cite[Eq. 10]{chen2004upper}:
\begin{equation} \label{eq:ceo_optimal_sumrate}
r = \frac{1}{2} \log^+ \left[ \frac{\sigma_\theta^2}{D^\star} \left( \frac{D^\star L}{ D^\star L - \sigma^2 + D^\star \sigma^2 / \sigma_\theta^2 }\right)^L  \right].
\end{equation}
For the special case where $r = n$ and $L=n$, we have
\begin{equation} \label{eq:ceo_optimal_sumrate2}
n = \frac{1}{2} \log_2 \left[ \frac{\sigma_\theta^2}{D^\star} \left(\frac{ D^\star n }{D^\star n - \sigma^2 + D^\star \sigma^2/\sigma_\theta^2 }  \right)^n  \right].
\end{equation}
Consider the distributed encoding setting (iii) in the case where $f(x) = \Ncal(0,\sigma^2)$ and the prior on $\Theta$ is $\pi = \Ncal(0,\sigma_\theta^2)$. The Gaussian CEO problem of \cite{viswanathan1997quadratic} with a unit bitrate $r_1=\ldots = r_n =1$ at each terminal and blocklength $k=1$ reduces to our distributed setting (iii). Since $D^\star$ satisfying \eqref{eq:ceo_optimal_sumrate2} describes the MSE in the CEO setting under an optimal allocation of the sum-rate $r = n$ among $n$ encoders, it provides a lower bound to the minimal MSE in estimating $\theta$ in the distributed setting. By considering the limit $n\rightarrow \infty$ in \eqref{eq:ceo_optimal_sumrate2}, we see that 
\[
D^\star = \frac{ 4\sigma^2 }{3n + 4 \sigma^2 / \sigma_\theta^2 } + O(n^{-2}) =  \frac{4\sigma^2}{3n} + O(n^{-2}). 
\]
This implies Proposition~\ref{prop:CEO}. 
%We note that although the lower bound \eqref{eq:ceo_bound} was derived assuming the optimal allocation of $n$ bits per observation among the encoders, this bound cannot be tightened by considering the MSE in the CEO setting while enforcing the condition $r_1=\ldots = r_n = 1$. Indeed, an upper bound for the CEO MSE under the condition $r_1=\ldots = r_n = 1$ follows from \cite{KipnisRini2019}, and leads to
%\[
%D^\star \leq  \left( \frac{1}{\sigma_\theta^2} +  \frac{3n}{4\sigma^2 + \sigma_\theta^2} \right)^{-1}   =
%\frac{4 \sigma^2}{3n} +  \frac{\sigma_\theta^2}{3n} + O(n^{-2}),
%\]
%which is equivalent to \eqref{eq:ceo_bound} when $\sigma_\theta$ is small.
